---
title: "Econ 491 - Ames Housing: Other Models"
author: "Housing 4 (Part 1: Yuxin Li, Part 2: Yafei Luo, Part 3: Zixuan Zhang, Part 4: Lintong Yin)"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Data Preprocessing (Same as Benchmark)

We reuse the exact same data cleaning and encoding pipeline from the benchmark model. This ensures all model comparisons are on equal footing.

```{r load-and-clean}
library(readr)
library(dplyr)
library(tidyverse)
library(tidyr)
library(glmnet)
library(ggplot2)
library(stargazer)

train <- read_csv("train.csv")
```

## Global Cleaning

```{r global-cleaning}
# Drop columns where more than 90% are NA
train <- train[, colMeans(is.na(train)) < 0.9]

# Drop low-value or redundant variables identified across all parts
train <- train %>%
  select(-Street, -Utilities, -Condition2) %>%
  mutate(LandSlope_binary = ifelse(LandSlope == "Gtl", 0, 1)) %>%
  select(-LandSlope)

train <- train %>%
  select(-MSSubClass, -LotFrontage)

drop_vars <- c("RoofMatl", "Heating", "ExterCond", "BsmtCond",
               "Exterior2nd", "BsmtFinSF1", "BsmtUnfSF", "BsmtFinSF2", "BsmtFinType2")
train <- train %>% select(-all_of(intersect(drop_vars, names(train))))

vars_to_drop <- c("CentralAir", "Electrical", "Functional",
                   "LowQualFinSF", "KitchenAbvGr", "BsmtHalfBath")
train <- train %>% select(-all_of(intersect(vars_to_drop, names(train))))

drop_vars2 <- c("Fence", "GarageQual", "GarageCond", "3SsnPorch",
                 "PoolArea", "MiscVal", "ScreenPorch", "PoolQC", "MiscFeature")
train <- train %>% select(-all_of(intersect(drop_vars2, names(train))))
```

## Part 1: Variables 1--20

```{r part1-encoding}
# MSZoning
pos <- which(names(train) == "MSZoning")
msz_dummies <- data.frame(
  MSZoning_FV = ifelse(train$MSZoning == "FV", 1, 0),
  MSZoning_RH = ifelse(train$MSZoning == "RH", 1, 0),
  MSZoning_RL = ifelse(train$MSZoning == "RL", 1, 0),
  MSZoning_RM = ifelse(train$MSZoning == "RM", 1, 0)
)
train <- cbind(train[, 1:(pos-1), drop = FALSE], msz_dummies,
               train[, (pos+1):ncol(train), drop = FALSE])

# LotShape
pos <- which(names(train) == "LotShape")
lotshape_dummy <- data.frame(LotShape_Reg = ifelse(train$LotShape == "Reg", 1, 0))
train <- cbind(train[, 1:(pos-1), drop = FALSE], lotshape_dummy,
               train[, (pos+1):ncol(train), drop = FALSE])

# LandContour
pos <- which(names(train) == "LandContour")
land_dummy <- data.frame(LandContour_Lvl = ifelse(train$LandContour == "Lvl", 1, 0))
train <- cbind(train[, 1:(pos-1), drop = FALSE], land_dummy,
               train[, (pos+1):ncol(train), drop = FALSE])

# LotConfig
pos <- which(names(train) == "LotConfig")
lotconfig_dummies <- data.frame(
  LotConfig_Corner  = ifelse(train$LotConfig == "Corner", 1, 0),
  LotConfig_CulDSac = ifelse(train$LotConfig == "CulDSac", 1, 0),
  LotConfig_FR2     = ifelse(train$LotConfig == "FR2", 1, 0),
  LotConfig_FR3     = ifelse(train$LotConfig == "FR3", 1, 0)
)
train <- cbind(train[, 1:(pos-1), drop = FALSE], lotconfig_dummies,
               train[, (pos+1):ncol(train), drop = FALSE])

# Neighborhood
pos <- which(names(train) == "Neighborhood")
nbhd_rank <- aggregate(SalePrice ~ Neighborhood, data = train, FUN = median)
nbhd_rank <- nbhd_rank[order(nbhd_rank$SalePrice), ]
n_nbhd <- nrow(nbhd_rank)
low_set  <- nbhd_rank$Neighborhood[1:floor(n_nbhd/3)]
high_set <- nbhd_rank$Neighborhood[(ceiling(2*n_nbhd/3)):n_nbhd]
nbhd_dummies <- data.frame(
  Neighborhood_High = ifelse(train$Neighborhood %in% high_set, 1, 0),
  Neighborhood_Low  = ifelse(train$Neighborhood %in% low_set, 1, 0)
)
train <- cbind(train[, 1:(pos-1), drop = FALSE], nbhd_dummies,
               train[, (pos+1):ncol(train), drop = FALSE])

# Condition1
pos <- which(names(train) == "Condition1")
positive_set <- c("PosA", "PosN")
negative_set <- c("Artery", "Feedr", "RRAn", "RRAe", "RRNe", "RRn")
condition_dummies <- data.frame(
  Condition_Pos = ifelse(train$Condition1 %in% positive_set, 1, 0),
  Condition_Neg = ifelse(train$Condition1 %in% negative_set, 1, 0)
)
train <- cbind(train[, 1:(pos-1), drop = FALSE], condition_dummies,
               train[, (pos+1):ncol(train), drop = FALSE])

# BldgType
pos <- which(names(train) == "BldgType")
bldg_dummies <- data.frame(
  BldgType_2fmCon = ifelse(train$BldgType == "2fmCon", 1, 0),
  BldgType_Duplex = ifelse(train$BldgType == "Duplex", 1, 0),
  BldgType_Twnhs  = ifelse(train$BldgType == "Twnhs", 1, 0),
  BldgType_TwnhsE = ifelse(train$BldgType == "TwnhsE", 1, 0)
)
train <- cbind(train[, 1:(pos-1), drop = FALSE], bldg_dummies,
               train[, (pos+1):ncol(train), drop = FALSE])

# HouseStyle
pos <- which(names(train) == "HouseStyle")
two_story_set <- c("2Story", "2.5Fin", "2.5Unf")
onehalf_set   <- c("1.5Fin", "1.5Unf")
split_set     <- c("SLvl", "SFoyer")
housestyle_dummies <- data.frame(
  HouseStyle_TwoStory = ifelse(train$HouseStyle %in% two_story_set, 1, 0),
  HouseStyle_OneHalf  = ifelse(train$HouseStyle %in% onehalf_set, 1, 0),
  HouseStyle_Split    = ifelse(train$HouseStyle %in% split_set, 1, 0)
)
train <- cbind(train[, 1:(pos-1), drop = FALSE], housestyle_dummies,
               train[, (pos+1):ncol(train), drop = FALSE])

# OverallQual & OverallCond: kept as-is

# YearBuilt & YearRemodAdd -> AgeBuilt & AgeRemod
pos_built <- which(names(train) == "YearBuilt")
AgeBuilt <- train$YrSold - train$YearBuilt
train <- cbind(train[, 1:(pos_built-1), drop = FALSE], AgeBuilt,
               train[, (pos_built+1):ncol(train), drop = FALSE])

pos_remod <- which(names(train) == "YearRemodAdd")
AgeRemod <- train$YrSold - train$YearRemodAdd
train <- cbind(train[, 1:(pos_remod-1), drop = FALSE], AgeRemod,
               train[, (pos_remod+1):ncol(train), drop = FALSE])
```

## Part 2: Variables 21--40

```{r part2-encoding}
# RoofStyle
train$RoofStyle <- as.factor(train$RoofStyle)
train$RoofStyle <- relevel(train$RoofStyle, ref = "Gable")

# Exterior1st
train <- train %>%
  mutate(Exterior1st = recode(Exterior1st,
    "VinylSd" = "VinylSd", "MetalSd" = "MetalSd", "HdBoard" = "HdBoard",
    "Wd Sdng" = "Wd Sdng", "Plywood" = "Plywood", "CemntBd" = "CemntBd",
    .default = "Other"))
train$Exterior1st <- as.factor(train$Exterior1st)
train$Exterior1st <- relevel(train$Exterior1st, ref = "VinylSd")

# MasVnrType
train <- train %>%
  mutate(MasVnrType = replace_na(MasVnrType, "None")) %>%
  mutate(MasVnrType = recode(MasVnrType,
    "BrkFace" = "BrkFace", "Stone" = "Stone", "None" = "None",
    "BrkCmn" = "None", .default = "None"))
train$MasVnrType <- as.factor(train$MasVnrType)
train$MasVnrType <- relevel(train$MasVnrType, ref = "None")

# ExterQual
qual_map <- c("Ex" = 5, "Gd" = 4, "TA" = 3, "Fa" = 2, "Po" = 1)
train$ExterQual <- qual_map[train$ExterQual]

# Foundation
train <- train %>%
  mutate(Foundation = recode(Foundation,
    "PConc" = "PConc", "CBlock" = "CBlock", "BrkTil" = "BrkTil",
    .default = "Other"))
train$Foundation <- as.factor(train$Foundation)
train$Foundation <- relevel(train$Foundation, ref = "PConc")

# BsmtQual
bsmt_qual_map <- c("Ex" = 5, "Gd" = 4, "TA" = 3, "Fa" = 2, "Po" = 1, "NA" = 0)
train$BsmtQual[is.na(train$BsmtQual)] <- "NA"
train$BsmtQual <- bsmt_qual_map[train$BsmtQual]

# BsmtExposure
expo_map <- c("Gd" = 3, "Av" = 2, "Mn" = 1, "No" = 0, "NA" = 0)
train$BsmtExposure[is.na(train$BsmtExposure)] <- "NA"
train$BsmtExposure <- expo_map[train$BsmtExposure]

# BsmtFinType1
fin_map <- c("GLQ" = 6, "ALQ" = 5, "BLQ" = 4, "Rec" = 3, "LwQ" = 2, "Unf" = 1, "NA" = 0)
train$BsmtFinType1[is.na(train$BsmtFinType1)] <- "NA"
train$BsmtFinType1 <- fin_map[train$BsmtFinType1]

# HeatingQC
qual_map <- c("Ex" = 5, "Gd" = 4, "TA" = 3, "Fa" = 2, "Po" = 1)
train$HeatingQC <- qual_map[train$HeatingQC]
```

## Part 3: Variables 41--60

```{r part3-encoding}
colnames(train)[colnames(train) == "1stFlrSF"] <- "X1stFlrSF"
colnames(train)[colnames(train) == "2ndFlrSF"] <- "X2ndFlrSF"

train$FireplaceQu <- as.character(train$FireplaceQu)
train$FireplaceQu[is.na(train$FireplaceQu)] <- "None"
train$GarageType <- as.character(train$GarageType)
train$GarageType[is.na(train$GarageType)] <- "None"
train$GarageFinish <- as.character(train$GarageFinish)
train$GarageFinish[is.na(train$GarageFinish)] <- "None"
train$KitchenQual <- as.character(train$KitchenQual)
train$KitchenQual[is.na(train$KitchenQual)] <- "None"

qual_map_fn <- function(x) {
  ifelse(x == "Ex", 5,
  ifelse(x == "Gd", 4,
  ifelse(x == "TA", 3,
  ifelse(x == "Fa", 2,
  ifelse(x == "Po", 1, 0)))))
}

train$KitchenQual_num  <- qual_map_fn(train$KitchenQual)
train$FireplaceQu_num  <- qual_map_fn(train$FireplaceQu)

train$GarageFinish_num <- ifelse(train$GarageFinish == "Fin", 3,
                          ifelse(train$GarageFinish == "RFn", 2,
                          ifelse(train$GarageFinish == "Unf", 1, 0)))

train$GarageYrBlt[is.na(train$GarageYrBlt)] <- median(train$GarageYrBlt, na.rm = TRUE)
train$GarageType_num <- as.integer(factor(train$GarageType))

train$KitchenQual  <- NULL
train$FireplaceQu  <- NULL
train$GarageFinish <- NULL
train$GarageType   <- NULL
```

## Part 4: Variables 61--78

```{r part4-encoding}
train <- train %>% select(-GarageArea, -MoSold, -YrSold)

train$GarageCars <- ifelse(train$GarageCars >= 4, 3, train$GarageCars)
train$GarageCars[is.na(train$GarageCars)] <- 0

paved_map <- c("N" = 0, "P" = 1, "Y" = 2)
train$PavedDrive <- paved_map[train$PavedDrive]

train$WoodDeckSF <- ifelse(
  train$WoodDeckSF + train$OpenPorchSF + train$EnclosedPorch > 0, 1, 0)
train <- train %>%
  rename(HasOutdoorSpace = WoodDeckSF) %>%
  select(-OpenPorchSF, -EnclosedPorch)

train$SaleType      <- ifelse(train$SaleType == "New", 1, 0)
train$SaleCondition <- ifelse(train$SaleCondition == "Normal", 1, 0)
```

## Final Preparation & Train-Test Split

```{r final-prep}
train$MasVnrArea[is.na(train$MasVnrArea)] <- median(train$MasVnrArea, na.rm = TRUE)

cat("Total columns after preprocessing:", ncol(train), "\n")
cat("Total rows:", nrow(train), "\n")

set.seed(123)
n <- nrow(train)
train_index <- sample(1:n, size = 0.8 * n)
train_in  <- train[train_index, ]
test_out  <- train[-train_index, ]

# Remove Id
train_in  <- train_in  %>% select(-Id)
test_out  <- test_out  %>% select(-Id)

cat("In-sample rows:", nrow(train_in), "\n")
cat("Out-of-sample rows:", nrow(test_out), "\n")
```

## Benchmark RMSE (Model 1)

```{r benchmark-rmse}
model_benchmark <- lm(SalePrice ~ ., data = train_in)
pred_benchmark  <- predict(model_benchmark, newdata = test_out)
rmse_benchmark  <- sqrt(mean((test_out$SalePrice - pred_benchmark)^2))
cat("Model 1 (Benchmark) RMSE:", round(rmse_benchmark, 2), "\n")
```


# Prepare Matrix Format for LASSO

`glmnet` requires numeric matrix input. `model.matrix()` automatically expands the 4 remaining factor columns (RoofStyle, Exterior1st, MasVnrType, Foundation) into dummy variables.

```{r matrix-prep}
# Ensure factor levels are consistent between train and test
for (col in names(train_in)) {
  if (is.factor(train_in[[col]])) {
    test_out[[col]] <- factor(test_out[[col]], levels = levels(train_in[[col]]))
  }
}

X_train <- model.matrix(SalePrice ~ ., data = train_in)[, -1]
y_train <- train_in$SalePrice

X_test  <- model.matrix(SalePrice ~ ., data = test_out)[, -1]
y_test  <- test_out$SalePrice

cat("Number of variables after dummy expansion:", ncol(X_train), "\n")
```


# Model 2: LASSO Variable Selection + OLS

LASSO (Least Absolute Shrinkage and Selection Operator) adds an L1 penalty to the OLS loss function, shrinking unimportant coefficients to exactly zero. We use cross-validation to select the optimal penalty parameter $\lambda$, then use the selected variables to re-estimate an OLS model for unbiased coefficients.

```{r model2-lasso}
set.seed(123)
cv_lasso <- cv.glmnet(X_train, y_train, alpha = 1)

plot(cv_lasso, main = "LASSO: Cross-Validation Error vs Lambda")
```

```{r model2-lambda}
cat("lambda.min:", cv_lasso$lambda.min, "\n")
cat("lambda.1se:", cv_lasso$lambda.1se, "\n")
```

## LASSO Selected Variables

We use `lambda.1se` for a more parsimonious model.

```{r model2-vars}
lasso_coef <- coef(cv_lasso, s = "lambda.1se")
lasso_vars <- rownames(lasso_coef)[which(lasso_coef != 0)]
lasso_vars <- lasso_vars[lasso_vars != "(Intercept)"]
cat("LASSO selected", length(lasso_vars), "variables:\n")
print(lasso_vars)
```

## OLS with LASSO-Selected Variables

```{r model2-ols}
formula_lasso <- as.formula(paste("SalePrice ~", paste(lasso_vars, collapse = " + ")))
model_lasso   <- lm(formula_lasso, data = train_in)
```

## Model 2 RMSE & Diagnostics

```{r model2-rmse}
pred_lasso <- predict(model_lasso, newdata = test_out)
rmse_lasso <- sqrt(mean((y_test - pred_lasso)^2))
cat("Model 2 (LASSO + OLS) RMSE:", round(rmse_lasso, 2), "\n")
```

```{r model2-stargazer, results='asis'}
stargazer(model_lasso, type = "latex",
          title = "Model 2: LASSO Selection + OLS",
          dep.var.labels = "Sale Price (USD)",
          omit.stat = c("f", "ser"), digits = 0, no.space = TRUE,
          header = FALSE)
```

```{r model2-plot}
ggplot(data.frame(Actual = y_test, Predicted = pred_lasso),
       aes(x = Actual, y = Predicted)) +
  geom_point(alpha = 0.4, color = "steelblue") +
  geom_abline(slope = 1, intercept = 0, color = "red",
              linetype = "dashed", linewidth = 1) +
  scale_x_continuous(labels = scales::comma) +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Model 2 (LASSO + OLS): Predicted vs Actual",
       subtitle = paste0("RMSE = $", format(round(rmse_lasso, 0), big.mark = ",")),
       x = "Actual Sale Price (USD)",
       y = "Predicted Sale Price (USD)") +
  theme_minimal(base_size = 13)
```


# Model 3: Log(SalePrice) + LASSO Selection + OLS

Following Wang (2025) and standard hedonic pricing literature, we apply a log transformation to SalePrice. This addresses the right-skewed distribution of housing prices and allows coefficients to be interpreted as approximate percentage effects.

```{r model3-lasso}
y_train_log <- log(y_train)

set.seed(123)
cv_lasso_log <- cv.glmnet(X_train, y_train_log, alpha = 1)
plot(cv_lasso_log, main = "Log Model LASSO: CV Error vs Lambda")
```

## Log LASSO Selected Variables

```{r model3-vars}
lasso_log_coef <- coef(cv_lasso_log, s = "lambda.1se")
lasso_log_vars <- rownames(lasso_log_coef)[which(lasso_log_coef != 0)]
lasso_log_vars <- lasso_log_vars[lasso_log_vars != "(Intercept)"]
cat("Log LASSO selected", length(lasso_log_vars), "variables:\n")
print(lasso_log_vars)
```

## OLS with Log-LASSO Variables

```{r model3-ols}
train_in$logSalePrice <- log(train_in$SalePrice)

formula_log <- as.formula(paste("logSalePrice ~", paste(lasso_log_vars, collapse = " + ")))
model_log   <- lm(formula_log, data = train_in)
```

## Model 3 RMSE & Diagnostics

Note: predictions are converted back to the original dollar scale via `exp()` before computing RMSE.

```{r model3-rmse}
pred_log <- exp(predict(model_log, newdata = test_out))
rmse_log <- sqrt(mean((y_test - pred_log)^2))
cat("Model 3 (Log + LASSO + OLS) RMSE:", round(rmse_log, 2), "\n")
```

```{r model3-stargazer, results='asis'}
stargazer(model_log, type = "latex",
          title = "Model 3: Log(SalePrice) + LASSO + OLS",
          dep.var.labels = "Log(Sale Price)",
          omit.stat = c("f", "ser"), digits = 4, no.space = TRUE,
          header = FALSE)
```

```{r model3-plot}
ggplot(data.frame(Actual = y_test, Predicted = pred_log),
       aes(x = Actual, y = Predicted)) +
  geom_point(alpha = 0.4, color = "steelblue") +
  geom_abline(slope = 1, intercept = 0, color = "red",
              linetype = "dashed", linewidth = 1) +
  scale_x_continuous(labels = scales::comma) +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Model 3 (Log + LASSO + OLS): Predicted vs Actual",
       subtitle = paste0("RMSE = $", format(round(rmse_log, 0), big.mark = ",")),
       x = "Actual Sale Price (USD)",
       y = "Predicted Sale Price (USD)") +
  theme_minimal(base_size = 13)
```


# Model 4: Log(SalePrice) + LASSO + Interaction Terms + OLS

Following Wang (2025), who included interaction terms where appropriate, and Zietz et al. (2008), who demonstrated that housing characteristics have varying effects across price levels, we add interaction terms to capture multiplicative effects between key variables.

## Verify Key Variables in LASSO Selection

```{r model4-check}
cat("OverallQual in selected?",      "OverallQual"      %in% lasso_log_vars, "\n")
cat("GrLivArea in selected?",        "GrLivArea"        %in% lasso_log_vars, "\n")
cat("AgeBuilt in selected?",         "AgeBuilt"         %in% lasso_log_vars, "\n")
cat("Neighborhood_High in selected?", "Neighborhood_High" %in% lasso_log_vars, "\n")
```

## Interaction Terms

We include three interaction terms based on economic reasoning:

1. **OverallQual $\times$ GrLivArea**: High-quality large homes command a disproportionate premium --- the value of each additional square foot is greater in a well-built home.
2. **OverallQual $\times$ AgeBuilt**: Higher quality construction retains value better over time, buffering the depreciation effect of age.
3. **Neighborhood\_High $\times$ GrLivArea**: Each additional square foot is worth more in a premium neighborhood than in a lower-tier location.

```{r model4-ols}
interaction_terms <- c("OverallQual:GrLivArea",
                       "OverallQual:AgeBuilt",
                       "Neighborhood_High:GrLivArea")

formula_interact <- as.formula(
  paste("logSalePrice ~",
        paste(lasso_log_vars, collapse = " + "),
        "+",
        paste(interaction_terms, collapse = " + "))
)

model_interact <- lm(formula_interact, data = train_in)
```

## Model 4 RMSE & Diagnostics

```{r model4-rmse}
pred_interact <- exp(predict(model_interact, newdata = test_out))
rmse_interact <- sqrt(mean((y_test - pred_interact)^2))
cat("Model 4 (Log + LASSO + Interaction + OLS) RMSE:", round(rmse_interact, 2), "\n")
```

```{r model4-stargazer, results='asis'}
stargazer(model_interact, type = "latex",
          title = "Model 4: Log + LASSO + Interaction + OLS",
          dep.var.labels = "Log(Sale Price)",
          omit.stat = c("f", "ser"), digits = 4, no.space = TRUE,
          header = FALSE)
```

```{r model4-plot}
ggplot(data.frame(Actual = y_test, Predicted = pred_interact),
       aes(x = Actual, y = Predicted)) +
  geom_point(alpha = 0.4, color = "steelblue") +
  geom_abline(slope = 1, intercept = 0, color = "red",
              linetype = "dashed", linewidth = 1) +
  scale_x_continuous(labels = scales::comma) +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Model 4 (Log + LASSO + Interaction): Predicted vs Actual",
       subtitle = paste0("RMSE = $", format(round(rmse_interact, 0), big.mark = ",")),
       x = "Actual Sale Price (USD)",
       y = "Predicted Sale Price (USD)") +
  theme_minimal(base_size = 13)
```


# Model Comparison

```{r comparison}
results <- data.frame(
  Model = c("Model 1: Benchmark (All Vars OLS)",
            "Model 2: LASSO Selection + OLS",
            "Model 3: Log + LASSO + OLS",
            "Model 4: Log + LASSO + Interaction + OLS"),
  Num_Vars = c(ncol(X_train),
               length(lasso_vars),
               length(lasso_log_vars),
               length(lasso_log_vars) + length(interaction_terms)),
  RMSE = c(rmse_benchmark,
           rmse_lasso,
           rmse_log,
           rmse_interact)
)

results <- results %>% arrange(RMSE)
print(results)
```

```{r comparison-plot}
ggplot(results, aes(x = reorder(Model, RMSE), y = RMSE)) +
  geom_col(fill = "steelblue", width = 0.6) +
  geom_text(aes(label = paste0("$", format(round(RMSE, 0), big.mark = ","))),
            hjust = -0.1, size = 3.5) +
  coord_flip() +
  scale_y_continuous(labels = scales::comma,
                     expand = expansion(mult = c(0, 0.2))) +
  labs(title = "Model Comparison: Out-of-Sample RMSE",
       x = NULL, y = "RMSE (USD)") +
  theme_minimal(base_size = 13)
```
